Geospatial Data Science - Final project

### 2025-03-18 final/final_wo_PostGIS.Rmd
Added a streamlined version

>>>>

## Regressions

### 1. Panel Data Set

First, we set-up a panel data set for our regressions.
```{r}
# Create road segment panel
# Step 1: Define time periods (monthly for 2021)
time_periods <- seq(from = as.Date("2021-01-01"), 
                    to = as.Date("2021-12-01"), 
                    by = "month")

# Step 2: Create road segments (either use existing or create grid cells)
# Option A: Use existing road segments from ca_roads
road_segments <- ca_roads %>%
  mutate(segment_id = row_number()) %>%
  select(segment_id, LINEARID, FULLNAME, geometry)

# Step 3: Generate panel by creating all combinations of segments and time periods
panel_data <- expand.grid(
  segment_id = road_segments$segment_id,
  month = time_periods
) %>%
  as_tibble()

# Step 4: Join spatial information
panel_data <- panel_data %>%
  left_join(road_segments %>% st_set_geometry(NULL) %>% select(segment_id), 
            by = "segment_id")
```

### 2. Treatment variables

We identify which segments had construction during which periods:
```{r}
# Identify construction treatment status for each segment-month
construction_panel <- df.const.sf %>%
  st_transform(st_crs(road_segments)) %>%
  # Create month variable
  mutate(month = floor_date(Start_Time, "month")) %>%
  # Spatial join with road segments
  st_join(road_segments) %>%
  # Group by segment and month
  group_by(segment_id, month) %>%
  # Create treatment indicator
  summarize(
    has_construction = 1,
    construction_count = n(),
    .groups = "drop"
  ) %>%
  st_set_geometry(NULL)

# Join treatment to panel
panel_data <- panel_data %>%
  left_join(construction_panel, by = c("segment_id", "month")) %>%
  mutate(has_construction = ifelse(is.na(has_construction), 0, 1))
```

### 3. Outcome variables

We count accidents by segment and time period:

```{r}
# Count accidents by segment and month
accident_counts <- df.acc.sf %>%
  st_transform(st_crs(road_segments)) %>%
  # Create month variable
  mutate(month = floor_date(Start_Time, "month")) %>%
  # Spatial join with road segments 
  st_join(road_segments) %>%
  # Group by segment and month
  group_by(segment_id, month) %>%
  # Count accidents and summarize severity
  summarize(
    accident_count = n(),
    severe_accident_count = sum(Severity >= 3, na.rm = TRUE),
    avg_severity = mean(Severity, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  st_set_geometry(NULL)

# Join accidents to panel
panel_data <- panel_data %>%
  left_join(accident_counts, by = c("segment_id", "month")) %>%
  mutate(
    accident_count = replace_na(accident_count, 0),
    severe_accident_count = replace_na(severe_accident_count, 0),
    avg_severity = replace_na(avg_severity, 0)
  )
```

### 4. Feature Engineering

We add several control variables:

```{r}
# 1. Add urban/rural classification (crucial for your research question)
# Get Census urban areas
urban_areas <- urban_areas(year = 2021)

# Identify if road segments are in urban areas
road_segments_urban <- road_segments %>%
  st_transform(st_crs(urban_areas)) %>%
  st_join(urban_areas %>% select(UACE, NAME, UATYP), join = st_intersects) %>%
  mutate(
    is_urban = !is.na(UACE),
    urban_type = ifelse(is.na(UATYP), "Rural", UATYP),
    urban_name = NAME
  ) %>%
  select(segment_id, is_urban, urban_type, urban_name)

# Join urban classification to panel
panel_data <- panel_data %>%
  left_join(road_segments_urban %>% st_set_geometry(NULL) %>% 
              select(segment_id, is_urban, urban_type), 
            by = "segment_id")

# 2. Add segment characteristics from road data
segment_chars <- road_segments %>%
  mutate(
    road_type = case_when(
      grepl("Interstate", RTTYP) ~ "Interstate",
      grepl("U.S. Highway", RTTYP) ~ "US Highway",
      grepl("State Highway", RTTYP) ~ "State Highway",
      TRUE ~ "Other"
    ),
    segment_length_km = as.numeric(st_length(geometry))/1000
  ) %>%
  select(segment_id, road_type, segment_length_km) %>%
  st_set_geometry(NULL)

# Join segment characteristics
panel_data <- panel_data %>%
  left_join(segment_chars, by = "segment_id")

# 3. Add time-based features
panel_data <- panel_data %>%
  mutate(
    year = year(month),
    month_num = month(month),
    quarter = quarter(month),
    is_holiday_month = month_num %in% c(7, 11, 12),  # July, Nov, Dec
    is_winter = month_num %in% c(12, 1, 2),
    is_summer = month_num %in% c(6, 7, 8)
  )

# 4. Normalize accident counts by segment length
panel_data <- panel_data %>%
  mutate(
    accidents_per_km = accident_count / segment_length_km,
    severe_accidents_per_km = severe_accident_count / segment_length_km
  )
```

And finally we also add the weather data from the initial accidents

```{r}
# Extract weather data from accidents dataset
weather_by_segment_month <- df.acc.sf %>%
  st_transform(st_crs(road_segments)) %>%
  mutate(month = floor_date(Start_Time, "month")) %>%
  st_join(road_segments) %>%
  group_by(segment_id, month) %>%
  summarize(
    avg_temperature = mean(Temperature, na.rm = TRUE),
    avg_precipitation = mean(Precipitation, na.rm = TRUE),
    pct_adverse_weather = mean(Weather_Condition %in% 
                              c("Rain", "Snow", "Fog", "Thunderstorm"), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  st_set_geometry(NULL)

# Join weather data
panel_data <- panel_data %>%
  left_join(weather_by_segment_month, by = c("segment_id", "month"))
```

### 5. Regression time

Now we can run the regression models based on our DiD design

```{r}
library(fixest) # For efficient fixed effects estimation

# 1. Base DiD Model
model1 <- feols(
  accident_count ~ has_construction | segment_id + month, 
  data = panel_data
)

# 2. Extended Model with Urban/Rural Heterogeneity
model2 <- feols(
  accident_count ~ has_construction + has_construction:is_urban | segment_id + month, 
  data = panel_data
)

# 3. Full Model with additional controls
model3 <- feols(
  accident_count ~ has_construction + has_construction:is_urban + 
                  has_construction:is_winter + has_construction:is_summer +
                  avg_temperature + avg_precipitation + pct_adverse_weather | 
                  segment_id + month,
  data = panel_data
)

# 4. Accident rate model (normalized by segment length)
model4 <- feols(
  accidents_per_km ~ has_construction + has_construction:is_urban | segment_id + month, 
  data = panel_data
)

# 5. Severity model
model5 <- feols(
  avg_severity ~ has_construction + has_construction:is_urban | segment_id + month, 
  data = panel_data
)

# Display results in a regression table
library(modelsummary)
modelsummary(list(model1, model2, model3, model4, model5),
             stars = TRUE,
             title = "DiD Estimates of Construction Impacts on Accidents")
```

### 6. Robustness Checks

We can also run some robustness checks to test the sensitivity of our results

```{r}
### 6. Robustness Checks

# 1. EVENT STUDY APPROACH - Analyze dynamic effects
# Create relative time variables to construction start
panel_data_event <- panel_data %>%
  group_by(segment_id) %>%
  # Find first month with construction for each segment
  mutate(
    first_construction = if(any(has_construction == 1)) min(month[has_construction == 1]) else NA,
    # Calculate months relative to construction start
    rel_month = as.numeric(difftime(month, first_construction, units = "days")) / 30,
    # Create factor for event study
    rel_period = case_when(
      is.na(rel_month) ~ NA_character_,
      rel_month <= -3 ~ "t-3+",
      rel_month <= -2 ~ "t-2",
      rel_month <= -1 ~ "t-1",
      rel_month < 0 ~ "t-0",
      rel_month < 1 ~ "t+0",
      rel_month < 2 ~ "t+1",
      rel_month < 3 ~ "t+2",
      TRUE ~ "t+3+"
    )
  ) %>%
  ungroup() %>%
  # Create dummy variables for each relative period
  mutate(
    rel_period = factor(rel_period, 
                        levels = c("t-3+", "t-2", "t-1", "t-0", "t+0", "t+1", "t+2", "t+3+"))
  )

# Run event study regression
event_model <- feols(
  accident_count ~ i(rel_period, ref = "t-1") | segment_id + month, 
  data = panel_data_event %>% filter(!is.na(rel_period))
)

# Visualize event study results
coef_plot <- iplot(event_model, main = "Event Study: Effect of Construction on Accidents",
      xlab = "Months Relative to Construction Start", ylab = "Effect on Accident Count")

# 2. PLACEBO TESTS - Randomly assign treatment
set.seed(123)
placebo_data <- panel_data %>%
  # Randomly shuffle the has_construction variable
  mutate(placebo_construction = sample(has_construction))

# Run placebo model
placebo_model <- feols(
  accident_count ~ placebo_construction | segment_id + month, 
  data = placebo_data
)

# Run multiple placebo tests
placebo_results <- data.frame(iteration = 1:100, estimate = NA, std.error = NA)
for(i in 1:100) {
  set.seed(i)
  # Create new placebo treatment
  temp_data <- panel_data %>%
    mutate(placebo_construction = sample(has_construction))

  # Run model
  temp_model <- feols(
    accident_count ~ placebo_construction | segment_id + month, 
    data = temp_data
  )

  # Store results
  placebo_results$estimate[i] <- coef(temp_model)[1]
  placebo_results$std.error[i] <- se(temp_model)[1]
}

# Plot placebo distribution vs. actual effect
placebo_plot <- ggplot(placebo_results, aes(x = estimate)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "black") +
  geom_vline(xintercept = coef(model1)[1], color = "red", linewidth = 1.5) +
  theme_minimal() +
  labs(title = "Distribution of Placebo Effects vs. Actual Effect",
       x = "Estimated Effect", y = "Count",
       caption = "Red line shows actual estimated effect")

# 3. BUFFER DISTANCE SENSITIVITY - Test different buffer definitions
# This requires going back to your spatial join code and creating different versions
# Here's a simplified approach (assuming you can regenerate panel data with different buffers)
buffer_results <- data.frame(
  buffer_dist = c(200, 300, 400, 500),
  estimate = c(0.12, 0.15, 0.11, 0.09),  # Example values - replace with actual results
  lower_ci = c(0.05, 0.08, 0.04, 0.01),  # from running your models with different buffer distances
  upper_ci = c(0.19, 0.22, 0.18, 0.17)
)

# Plot buffer sensitivity results
buffer_plot <- ggplot(buffer_results, aes(x = factor(buffer_dist), y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.2) +
  theme_minimal() +
  labs(title = "Sensitivity to Buffer Distance",
       x = "Buffer Distance (meters)", y = "Estimated Effect on Accident Count")

# 4. ALTERNATIVE FIXED EFFECTS - Test different specifications
alt_fe_model1 <- feols(accident_count ~ has_construction, data = panel_data)  # No FE
alt_fe_model2 <- feols(accident_count ~ has_construction | segment_id, data = panel_data) # Only segment FE
alt_fe_model3 <- feols(accident_count ~ has_construction | month, data = panel_data) # Only time FE
alt_fe_model4 <- feols(accident_count ~ has_construction | segment_id + month, data = panel_data) # Both FE (baseline)

# Compare models
fe_comparison <- modelsummary(list("No FE" = alt_fe_model1, 
                                  "Segment FE" = alt_fe_model2,
                                  "Time FE" = alt_fe_model3,
                                  "Both FE" = alt_fe_model4),
                            stars = TRUE)

# 5. DIFFERENT OUTCOME TRANSFORMATIONS - Test sensitivity to functional form
transform_model1 <- feols(accident_count ~ has_construction | segment_id + month, 
                       data = panel_data) # Linear (baseline)
transform_model2 <- feols(log(accident_count + 1) ~ has_construction | segment_id + month, 
                       data = panel_data) # Log transform
transform_model3 <- fepois(accident_count ~ has_construction | segment_id + month, 
                       data = panel_data) # Poisson

# Compare transformations
transform_comparison <- modelsummary(list("Linear" = transform_model1,
                                        "Log" = transform_model2,
                                        "Poisson" = transform_model3),
                                   stars = TRUE)

# 6. Combine robustness visualizations into a panel
library(patchwork)
robustness_panel <- (coef_plot / placebo_plot) | (buffer_plot / fe_comparison)
robustness_panel + plot_annotation(
  title = "Robustness Checks for Construction Safety Analysis",
  theme = theme(plot.title = element_text(size = 16, hjust = 0.5))
)
```

Additional validation, propensity score matching:

```{r}
library(MatchIt)
matched_segments <- matchit(
  has_construction ~ road_type + segment_length_km + is_urban,
  data = panel_data %>% filter(month == min(month)),
  method = "nearest"
)
```
